Subset coercions in \Coq \\ \small{Draft - \today}

### 2006-08-23 papers/article/main.tex
Little remaining fixes

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}
\usepackage{alltt}

\def\text{\mbox}

%\usepackage{amsmath}
%\usepackage{amsthm}

@@

\newenvironment{code}{\begin{alltt}\vspace{-1em}}{\end{alltt}\vspace{-1em}}

@@

\author{Matthieu Sozeau\inst{1}}

\institute{LRI, Paris Sud XI University\\\email{sozeau@lri.fr}}
\title{Subset coercions in \Coq}
\date{\today}

@@

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a pratical functionnal programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgments can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term. A prototype implementation of this process has been
  implemented.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent complex types, we immediatly get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{head} which returns
the first element of a list. In \ML{}:
\begin{code}
  let head = function hd :: tl -> hd | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
Definition head (l : list A) : option A := 
  match l with
    hd :: tl -> Some hd
  | [] -> None
  end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, using a
richer specification: 
\begin{code}
Definition head (l : list A) : forall p : l <> [] -> A.
\end{code} 
Now the definition's body requires some heavy plumbing of the code which is 
not afordable when programming. We propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is a weak, \emph{decidable} type-checking
procedure along with an elaboration into partial \Coq terms. Using
\Russell, we would write:
\begin{code}
  Definition head \{ l : list A | l <> [] \} : A := 
  match l with
    hd :: tl -> hd
  | [] -> _
  end
\end{code}

The specification shows that we're definining a partial function yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proof in the second branch. After type-checking,
the proof assistant will ask the user to prove that in the context where
\id{l} is a non-empty list and $l = []$, we can create an object of type
$A$ (which should be obvious as the context is contradictory).

@@

The \Program{} tactic by C. Parent \cite{conf/mpc/Parent95} had the same
goal as ours but a slightly different method for achieving it. It was strongly
linked to the extraction mechanism included in \Coq, both theoreticaly
and practically. Sketching the mechanism, she defined a weakened
extraction operation on \CIC{} terms which could be inverted because it left enough
information in the extracted term to rebuild a partial proof. The
mechanism, while general and theoreticaly well thought out, required
some heuristics and did not integrate smoothly with the \Coq{}
environment. In particular it lacked the pervasivness our method has,
being applicable in a wide variety of situations in the proof assistant
environment.

Instead of trying to find a general method for synthetising proofs from
programs in the \CICfull, we have integrated a method which permits to
link strong specifications and purely algorithmical code. This method,
known as \ps{} in the \PVS{} system has been used with great success and
fits naturally with how we write specifications using the subset type \cite{progmartinlof}.
The main contribution of this paper is to show how the \ps{} method 
can be adapted in a proof assistant with proof terms, which formally 
justifies the extension in the first place.

The remainder of the article is organized as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of traductions between
judgements of the two systems in section \ref{sec:Traduction}. Then, we present a prototype
implementation with some case studies, and finally we give our
conclusions on this work and discuss future directions.

@@

\begin{figure}
  \begin{center}
    \def\type{\typed}
    \AXC{$`G \type A : \Set$}
    \AXC{$`G, x : A \type P : \Prop$}
    \BIC{$`G \type \mysubset{x}{A}{P} : \Set$}
    \DP

@@

\subsection{Typing and conversion}
% Our presentation of \ps{} differs slightly from the original one
% essentially because of the proof terms (hence the use of a judgement
% instead of a function for checking subtyping), but we have also extended
% it by . 
\subsubsection{Type system}
\begin{figure}[t]
  \begin{center}
    \def\infvspace{0.1in}
    \def\seq{\typed}
    \def\fCenter{\wf}

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{-2em}
  \caption{\CIC typing judgement}
  \label{fig:cic-decl-rules}
  \vspace{-1em}
\end{figure}

@@

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.  In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgement $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $(s_1, s_2, s_2)$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ _ : unit `| A \}, B$ products to benefit from
our languages subtyping mechanism.

@@

Likewise, we need to restrict the system so that it becomes property
irrelevant for the propositions appearing in subset types. Indeed, in
\PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgment $0 : \mysubset{x}{`N}{P}$, but
in a classical Calculus of Constructions with sigma types or inductive
types, the subset type is a first class object defined with proper
introduction and elimination terms (figure \vref{fig:cci-subset}). What
we need to do is forbid the use of these terms in \Russell{} terms,
leaving only the type former (figure \vref{fig:typing-decl-rules}).

@@

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgement.
The judgment $`G \typed T \subd{} U : s$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is equivalent to $U$ in
environment $`G$, both being sorted by $s$.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

\subsubsection{Properties}
We have proved the metatheory for this system in \Coq{}
\cite{sozeau:Coq/Russell/meta-web}. We have assumed that it is strongly
normalizing (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR doesn't depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we prefered to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalization could also be proved, perhaps more easily,
using the generic proof method developed in \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. 
% This notion is presented in \PVS{} by
% means of a $\mu_0$ operator which was readily adaptable in our setting
% (figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
% new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
% $\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
\end{proof}
\end{paragraph}

@@

Finaly, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U~s, `G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}
\begin{proof}
  Rules are syntax-directed.
\end{proof}

@@

\section{From \lng{} to \CICq{}}
\label{sec:Traduction}
We will now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CIC
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

Our interpretation will build a full-fledged \CIC term from its
algorithmic skeleton and a rich type written in \Russell. Obviously, we
cannot infer proof terms where they are needed in \CIC, but we can build
a \emph{partial} term, leaving typed holes where proofs are
needed. Hence we add a rule to build existential variables (or
metavariables) in the target calculus ($\typeq$ denotes the new system's typing judgement):
\begin{prooftree}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
\end{prooftree}

@@

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which build a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
improved readability. The judgement $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{`G}{T}$ to an object of type $\ip{`G}{U}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed $\Coq$ term. We
could also directly use the interpreted types but that complicates the
proof.

@@

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instanciation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied. Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and well suited for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

@@

Next, we have the rules for products and sums, which compose
coercions. In the \irule{SubProd} rule, we first coerce the argument to
an object of type $\ip{T}{`G}$. Then we apply it to the coerced
function, resulting in an object $\ctxdot~(c_1[x])$ of type
$\ip{V}{`G, x : T}[c_1[x]/x]$. We'll see later that this type is equivalent to
$\ip{V}{`G, x : U}$, so we can apply the second coercion and get an
object of type $\ip{W}{`G}$. There is a little twist when coercing the
second component of a dependent sum. Indeed, we must simultaneously
substitute the first projection of the coerced sum in the second
coercion which was typed in environment $`G, x : T$ and also instanciate
the coercion with the second component. Clearly, the $\ctxdot$ of the
former shouldn't be instanciated with the later, so we must first
instanciate $c_2$ and then substitute $\pi_1~\ctxdot$ for $x$ 
($x$ cannot appear in $\pi_2~t$ where $t : \Sigma x : T.V$).

@@

\interpfigs[Interpretion of terms]

@@

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms to a system with only
$\beta\pi$-equivalence. 
Indeed, if our algorithmic system has transitivity and symetry of
coercions, we expect that the derived coercions will inherit these
properties, for example:
\begin{proposition}[Symetry of coercions]
  If $\subimpl{`G}{c}{T}{U}{s}$ then there exists $c^{-1}$ so that
  $\subimpl{`G}{c^{-1}}{U}{T}{s}$ and $c `o c^{-1} `= c^{-1} `o c `=
  \ctxdot$.
\end{proposition}

>>>>

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}
\usepackage{alltt}
\usepackage{color}
\def\text{\mbox}
\pagestyle{plain}
%\usepackage{amsmath}
%\usepackage{amsthm}

@@

\newenvironment{code}{\begin{alltt}\vspace{-1em}}{\vspace{-1em}\end{alltt}}
% Caml2Tex

\definecolor{KWLanguage_color}{rgb}{0.6,0,0}
\definecolor{KWStructure_color}{rgb}{0.8,0.8,0}
\definecolor{KWModule_color}{rgb}{0.7,0.5,0}
\definecolor{KWSubprog_color}{rgb}{0.6,0,0.6}
\definecolor{KWException_color}{rgb}{0.8,0,0}
\definecolor{Constructor_color}{rgb}{0,0.5,0}
\definecolor{Comment_color}{rgb}{0.6,0.6,0.6}
\definecolor{String_color}{rgb}{0.6,0.5,0.5}
\definecolor{Varfunction_color}{rgb}{0.2,0.2,0.8}
\definecolor{Constant_color}{rgb}{0.6,0.5,0.5}
\definecolor{Variable_color}{rgb}{0.2,0.2,0.8}

\newcommand\KWLanguage[1]{\rmfamily{\color{KWLanguage_color}{#1}}}
\newcommand\KWStructure[1]{\sffamily{\color{KWStructure_color}{#1}}}
\newcommand\KWModule[1]{\rmfamily{\color{KWModule_color}{#1}}}
\newcommand\KWSubprog[1]{\rmfamily{\color{KWSubprog_color}{#1}}}
\newcommand\KWException[1]{\rmfamily{\color{KWException_color}{#1}}}
\newcommand\Constructor[1]{\textbf{\color{Constructor_color}{#1}}}
\newcommand\Comment[1]{\rmfamily{\color{Comment_color}{#1}}}
\newcommand\String[1]{\rmfamily{\color{String_color}{#1}}}
\newcommand\Varfunction[1]{\sffamily{\color{Varfunction_color}{#1}}}
\newcommand\AVariable[1]{\textit{\color{Variable_color}{#1}}}
\newcommand\Constant[1]{\texttt{\color{Constant_color}{#1}}}
\newcommand\pow{$^{\wedge}$}
\newcommand\bs{$\backslash$}

@@

\author{Matthieu Sozeau}

\institute{LRI, Paris Sud XI University\\\email{sozeau@lri.fr}}
\title{Subset coercions in \Coq \\ \small{Draft - \today}}
\date{\today}

@@

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a practical functional programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgments can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term. A prototype implementation of this process has been
  implemented.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{head} which returns
the first element of a list. In \ML{}:
\begin{code}
  let head = function hd :: tl -> hd | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
  Definition head (l : list A) : option A :=
    match l with
      hd :: tl => Some hd
    | [] => None
    end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, using a
richer specification: 
\begin{code}
  Definition head (l : list A) : forall p : l <> [] -> A.
\end{code} 
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. We propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is a weak, \emph{decidable} type-checking
procedure along with an elaboration into partial \Coq terms. Using
\Russell, we would write:
\begin{code}
  Definition head \{ l : list A | l <> [] \} : A := 
  match l with
    hd :: tl -> hd
  | [] -> _
  end
\end{code}

The specification shows that we're defining a partial function yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proof in the second branch. After type-checking,
the proof assistant will ask the user to prove that in the context where
\id{l} is a non-empty list and $l = []$, we can create an object of type
$A$ (which should be obvious as the context is contradictory).

@@

The \Program{} tactic by C. Parent \cite{conf/mpc/Parent95} had the same
goal as ours but a slightly different method for achieving it. It was strongly
linked to the extraction mechanism included in \Coq, both theoretically
and practically. Sketching the mechanism, she defined a weakened
extraction operation on \CIC{} terms which could be inverted because it left enough
information in the extracted term to rebuild a partial proof. The
mechanism, while general and theoretically well thought out, required
some heuristics and did not integrate smoothly with the \Coq{}
environment. In particular it lacked the pervasiveness our method has,
being applicable in a wide variety of situations in the proof assistant
environment.

Instead of trying to find a general method for synthetizing proofs from
programs in the \CICfull, we have integrated a method which permits to
link strong specifications and purely algorithmical code. This method,
known as \ps{} in the \PVS{} system has been used with great success and
fits naturally with how we write specifications using the subset type \cite{progmartinlof}.
The main contribution of this paper is to show how the \ps{} method 
can be adapted in a proof assistant with proof terms, which formally 
justifies the extension in the first place.

The remainder of the article is organized as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of translations between
judgments of the two systems in section \ref{sec:Translation}. Then, we present a prototype
implementation in section \ref{sec:Program}, and finally we give our
conclusions on this work and discuss future directions in section \ref{sec:Conclusion}.

@@

\begin{figure}[h]
  \begin{center}
    \def\type{\typed}
    \AXC{$`G \type A : \Set$}
    \AXC{$`G, x : A \type P : \Prop$}
    \BIC{$`G \type \mysubset{x}{A}{P} : \Set$}
    \DP

@@

\subsection{A weaker type system}
To formalize this idea in \Coq, we simply weaken the type system so that
it doesn't require the terms to contain the proof components for objects
of subset types. This permits to have a simple language for code while
retaining the richness of \Coq's specification language. Once we have a
derivation in this new type system, we can translate it to a partial
\Coq derivation, where the missing parts are represented by
metavariables. It can then be completed by instantiating these holes
with actual proofs.

\subsubsection{Type system}
\begin{figure}[h]
  \begin{center}
    \def\infvspace{0.1in}
    \def\seq{\typed}
    \def\fCenter{\wf}

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{-2em}
  \caption{\CIC typing judgment}
  \label{fig:cic-decl-rules}
  \vspace{-1em}
\end{figure}

@@

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgment $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.  In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgment $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $(s_1, s_2, s_2)$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ products to benefit from
the subtyping mechanism.

@@

Likewise, we need to restrict the system so that it becomes property
irrelevant for the propositions appearing in subset types. Indeed, in
\PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgment $0 : \mysubset{x}{`N}{x \neq 0}$, but
in a classical Calculus of Constructions with sigma types or inductive
types, the subset type is a first class object defined with proper
introduction and elimination terms (figure \vref{fig:cci-subset}). What
we need to do is forbid the use of these terms in \Russell{} terms,
leaving only the type former (figure \vref{fig:typing-decl-rules}).

@@

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgment.
The judgment $`G \typed T \subd{} U : s$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is equivalent to $U$ in
environment $`G$, both being sorted by $s$.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. We use a judgmental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgment constructs an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgments because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

\subsubsection{Properties}
We have proved the metatheory for this system in \Coq{}
\cite{sozeau:Coq/Russell/meta-web}. We have assumed that it is strongly
normalizing (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR doesn't depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalization could also be proved, perhaps more easily,
using the generic proof method developed in \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
Our algorithmic judgment $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. 
% This notion is presented in \PVS{} by
% means of a $\mu_0$ operator which was readily adaptable in our setting
% (figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
% new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
% $\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgment for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
\end{proof}
\end{paragraph}

@@

Finally, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U~s, `G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}
\begin{proof}
  Rules are syntax-directed.
\end{proof}

@@

\section{From \lng{} to \CICq{}}
\label{sec:Translation}
We will now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CIC
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

Our interpretation will build a full-fledged \CIC term from its
algorithmic skeleton and a rich type written in \Russell. Obviously, we
cannot infer proof terms where they are needed in \CIC, but we can build
a \emph{partial} term, leaving typed holes where proofs are
needed. Hence we add a rule to build existential variables (or
metavariables) in the target calculus ($\typeq$ denotes the new system's typing judgment):
\begin{prooftree}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
\end{prooftree}

@@

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgment which build a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
improved readability. The judgment $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed \Coq{} term. We
could also directly use the interpreted types but that complicates the
proof.

@@

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied. Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgment $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and well suited for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

@@

Next, we have the rules for products and sums, which compose
coercions. In the \irule{SubProd} rule, we first coerce the argument to
an object of type $\ip{T}{`G}$. Then we apply it to the coerced
function, resulting in an object $\ctxdot~(c_1[x])$ of type
$\ip{V}{`G, x : T}[c_1[x]/x]$. We'll see later that this type is equivalent to
$\ip{V}{`G, x : U}$, so we can apply the second coercion and get an
object of type $\ip{W}{`G}$. There is a little twist when coercing the
second component of a dependent sum. Indeed, we must simultaneously
substitute the first projection of the coerced sum in the second
coercion which was typed in environment $`G, x : T$ and also instantiate
the coercion with the second component. Clearly, the $\ctxdot$ of the
former shouldn't be instantiated with the later, so we must first
instantiate $c_2$ and then substitute $\pi_1~\ctxdot$ for $x$ 
($x$ cannot appear in $\pi_2~t$ where $t : \Sigma x : T.V$).

@@

\interpfigs[Interpretation of terms]

@@

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms into a system with only
$\beta\pi$-equivalence. 
Indeed, if our algorithmic system has transitivity and symmetry of
coercions, we expect that the derived coercions will inherit these
properties, for example:
\begin{proposition}[Symmetry of coercions]
  If $\subimpl{`G}{c}{T}{U}{s}$ then there exists $c^{-1}$ so that
  $\subimpl{`G}{c^{-1}}{U}{T}{s}$ and $c `o c^{-1} `= c^{-1} `o c `=
  \ctxdot$.
\end{proposition}

Now take the following example:
\begin{example}[Symmetry for products]
  Let $T, U = \Pi x : A.B$. The coercions $c, c^{-1}$ derived by
  the algorithm must have the form $\lambda x :
  \ip{A}{`G}.c_2[\ctxdot~c_1[x]]$, hence we must include the $\eta$
  rule in order to have $c `o c^{-1} `= \ctxdot$.
\end{example}
